---
title: "Course Project 1"
author: "Bryant Gagliardi"
date: "9/15/2018"
output: html_document
---
#Bryant Gagliardi Course Project 1

Read in the data
```{r setup, echo = TRUE}

#Set the working directory
setwd ("//prod.protected.ind/User/user06/bg8684/My Documents/Data_course/Project_1")
        #read in the data
        stepdata <- (read.csv("activity.csv", header=TRUE))
```


What is mean total number of steps taken per day?
```{r stepsperday, echo = TRUE}
# 1. Calculate the total number of steps taken per day:
#remove NAs for this part of the analysis
completesteps <- na.omit(stepdata)
#load the plyr package
library(plyr)
#summarise and sum the steps data by date, into a dataframe
dailysteps <- 
        ddply(completesteps, .(date), summarize, daily_total_steps = sum(steps))
        

#2. Make histogram of the number of steps taken each day:
#assign "number of steps per day" as a variable
daily_total_steps <- dailysteps$daily_total_steps
        #make a labelled histogram 
        hist(daily_total_steps, main = "Histogram of total steps taken per day",
        xlab = "Number of steps", ylab = "Frequency")

#3. Calculate and report the mean and median of the total number of steps taken 
#per day:
#take the mean of all daily step values in the "dailysteps" dataframe"
rmean <- mean(dailysteps[,"daily_total_steps"])
        #take the median of all daily step values
        rmedian <- median(dailysteps[,"daily_total_steps"])
```

The mean number of steps taken per day is `r rmean`, the median number of steps
is `r rmedian`.



What is the average daily activity pattern?

```{r activitypattern, echo=TRUE}

#1.Make a time series plot:
#summarise and average the steps data by interval, into a dataframe
intervalsteps <- ddply(completesteps, .(interval), summarize, 
                       mean_steps_per_interval = mean(steps))
        #assign the relevant variables for the plot
        int <- intervalsteps$interval
        ave_steps <- intervalsteps$mean_steps_per_interval
        #create the timeseries plot
        plot(int, ave_steps, type = "l", main = "Daily activity",
                xlab = "Interval", ylab = "Average no. steps")

#2. Which interval has the max average steps?       
        #determine which row in the dataset has the highest average steps
        maxinterval <- which(ave_steps==max(ave_steps))
        #determine which interval corresponds to this row
        moststeps <- intervalsteps[maxinterval,"interval"]
```
The most steps on average are taken in interval number `r moststeps`.


Imputing missing values
```{r missingvalues, echo=TRUE}
#1. Calculate and report number of missing values:
#count the number of steps "NA" values in the original dataset
        StepNAs <- sum(is.na(stepdata$steps))

#2. Devise a strategy for filling in NAs:
        #replace NAs with the average number of steps per interval
#3. Create a new dataset that is equal to the original but with missing data 
        #added:
#fill in NAs with the average number of steps per interval
stepdata[is.na(stepdata)] <- mean(completesteps$steps)
        
#4.Make a histogram of steps taken each day,
imputedsteps <- 
        ddply(stepdata, .(date), summarize, daily_total_steps = sum(steps))

        hist(imputedsteps$daily_total_steps, main = 
        "Histogram of total steps taken per day",
        xlab = "Number of steps", ylab = "Frequency")
        
#...calculate the mean and median total number of steps taken per day
       MnDTS <- mean(imputedsteps$daily_total_steps)
       MdDTS <- median(imputedsteps$daily_total_steps)
       meandiff <- MnDTS - rmean
       meddiff <- MdDTS - rmedian

```
1. There are `r StepNAs` NA values in the dataset
4. The mean total number of daily steps taken is `r MnDTS`, the median total
number of daily steps taken is `r MdDTS`. The difference between the mean
with imputed values and the mean without imputed values is `r meandiff`. The 
difference between the median with imputed values and the median without imputed 
values is `r meddiff`. 


Are there differences in activity patterns between weekdays and weekends?
```{r weekdaysweekends, echo=TRUE}
#1. Create a new factor variable in the dataset with two levels - 
#"weekday" and "weekend":
#for days containing capital S, assign them weekends, otherwise weekdays
stepdata$day = weekdays(as.Date(stepdata$date))
        stepdata$WD <- ifelse(grepl("S", stepdata$day, ignore.case=FALSE), 
                              "Weekend","Weekday")

#2. Make panel plot comparing activity for weekdays and weekends:
        #make separate weekday and weekend datasets
        wkdy <- stepdata[which(stepdata$WD == 'Weekday'),]
        wknd <- stepdata[which(stepdata$WD == 'Weekend'),]
        #average the steps per interval for each dataset
        wkdysteps <- ddply(wkdy, .(interval), summarize, 
                       mean_steps_per_interval = mean(steps))

        wkndsteps <- ddply(wknd, .(interval), summarize, 
                       mean_steps_per_interval = mean(steps))
        #plot the results for comparison
        par(mfrow=c(1,2))
        plot(x=wkdysteps$interval,y=wkdysteps$mean_steps_per_interval, 
             type="l", ylab = "Average no. steps", xlab = "Interval", 
             main = "Weekday activity")
        plot(wkndsteps$interval,wkndsteps$mean_steps_per_interval, 
             type="l",ylab="", xlab = "Interval", main="Weekend activity")
```
